---
title: "4.5 Computing Posteriors: From Grids to Stan"
subtitle: "Grid and Laplace approximation, Metropolis-Hastings, Gibbs sampling, Hamiltonian Monte Carlo, and a Bayesian multilevel model in Stan"
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(lme4)
source("R/player_helpers.R")

passes <- load_passes()
results <- readRDS("data/bayes_results.rds")
```

## Learning objectives

After this tutorial, you should be able to:

- explain why most posterior distributions must be computed numerically;
- implement grid approximation and the Laplace approximation for a two-parameter posterior;
- implement the Metropolis-Hastings algorithm and diagnose a chain with its acceptance rate, trace, autocorrelation, effective sample size, and R-hat;
- describe Gibbs sampling and implement it for the normal hierarchical model;
- explain the idea of Hamiltonian Monte Carlo and why it scales to many parameters;
- read a multilevel model written in Stan and interpret its output; and
- compare posterior summaries with `lme4` estimates and rank players with uncertainty.

## Why computation is needed

Bayes' rule gives the posterior up to a constant,

$$
p(\theta\mid y)=\frac{p(y\mid\theta)p(\theta)}{\int p(y\mid\theta)p(\theta)\,d\theta}.
$$

Outside conjugate families the integral in the denominator has no closed form, and neither do the quantities we report: posterior means, credible intervals, and probabilities such as $P(\theta_A>\theta_B\mid y)$. Every one of them is an integral over the posterior. The methods in this tutorial are different ways of computing such integrals.

## A two-parameter example

Take one passer and model completion as a function of throw depth:

$$
P(y_i=1)=\sigma(\beta_0+\beta_1x_i),
\qquad
\beta_0,\beta_1\sim N(0,\ 2.5^2),
$$

where $x_i$ is standardized air yards. With prior standard deviation 2.5 on the log-odds scale, the prior is weakly informative: it rules out absurd coefficients without favoring any plausible value.

```{r}
one_passer <- passes %>% filter(passer == "Patrick Mahomes")
X <- cbind(1, one_passer$air_yards_z)
y <- one_passer$complete
nrow(one_passer)

log_posterior <- function(beta) logistic_log_posterior(beta, X, y, prior_sd = 2.5)
```

### Grid approximation

Evaluate the unnormalized log posterior on a grid, exponentiate, and divide by the sum. The normalized grid is a discrete approximation to the posterior from which any summary can be computed.

```{r}
beta0_grid <- seq(0.3, 1.3, length.out = 121)
beta1_grid <- seq(-1.1, -0.2, length.out = 121)
grid <- expand.grid(beta0 = beta0_grid, beta1 = beta1_grid)
grid$log_post <- apply(grid, 1, function(row) log_posterior(c(row[["beta0"]], row[["beta1"]])))
grid$posterior <- exp(grid$log_post - max(grid$log_post))
grid$posterior <- grid$posterior / sum(grid$posterior)

grid_summary <- data.frame(
  parameter = c("beta0", "beta1"),
  mean = c(sum(grid$posterior * grid$beta0), sum(grid$posterior * grid$beta1)),
  sd = c(
    sqrt(sum(grid$posterior * grid$beta0^2) - sum(grid$posterior * grid$beta0)^2),
    sqrt(sum(grid$posterior * grid$beta1^2) - sum(grid$posterior * grid$beta1)^2)
  )
)
grid_summary %>% knitr::kable(digits = 3)
```

```{r}
#| fig-cap: "Posterior of the intercept and air-yards slope on a grid."
#| fig-alt: "Contour plot of a roughly elliptical posterior centered near an intercept of 0.8 and a slope of minus 0.6."
ggplot(grid, aes(beta0, beta1, z = posterior)) +
  geom_contour_filled(bins = 8) +
  labs(x = "Intercept", y = "Air-yards slope", fill = "Posterior") +
  theme_minimal() +
  theme(legend.position = "none")
```

A grid with 121 points per dimension needs $121^2$ evaluations here. With 10 parameters it would need $121^{10}$, which is why grids are limited to one or two dimensions.

### Laplace approximation

Near its mode a smooth log posterior is approximately quadratic, so the posterior is approximately normal:

$$
p(\theta\mid y)\approx N\!\left(\hat\theta,\ \left[-\nabla^2\log p(\hat\theta\mid y)\right]^{-1}\right),
$$

where $\hat\theta$ is the posterior mode and the covariance is the inverse of the negative Hessian at the mode. Finding the mode is an optimization problem, which is cheap in any dimension.

```{r}
laplace <- laplace_approximation(log_posterior, c(0, 0))

data.frame(
  parameter = c("beta0", "beta1"),
  mode = laplace$mode,
  sd = sqrt(diag(laplace$covariance))
) %>%
  knitr::kable(digits = 3)
```

The Laplace approximation matches the grid closely because this posterior is nearly normal. It is also the approximation `glmer()` uses for its marginal likelihood. It becomes poor for skewed posteriors such as those of variance parameters.

## Markov chain Monte Carlo

Monte Carlo replaces integrals with averages over draws: if $\theta^{(1)},\dots,\theta^{(S)}$ are draws from the posterior, then

$$
E[g(\theta)\mid y]\approx\frac{1}{S}\sum_{s=1}^{S}g\!\left(\theta^{(s)}\right).
$$

Markov chain Monte Carlo (MCMC) produces such draws without knowing the normalizing constant by constructing a Markov chain whose stationary distribution is the posterior. Successive draws are correlated, and early draws depend on the starting point, so a chain is run long enough to forget where it started (the warm-up) and then long enough to average out the correlation.

### Metropolis-Hastings

The random-walk Metropolis algorithm is:

1. from the current state $\theta$, propose $\theta^{*}=\theta+\epsilon$ with $\epsilon\sim N(\mathbf 0,s^2 I)$;
2. compute the ratio $r=\dfrac{p(\theta^{*}\mid y)}{p(\theta\mid y)}$, in which the normalizing constant cancels;
3. accept $\theta^{*}$ with probability $\min(1,r)$, otherwise keep $\theta$; and
4. repeat.

Every proposal that increases the posterior is accepted, and a proposal that decreases it is accepted with probability equal to the ratio. Over many steps the chain spends time in each region in proportion to its posterior probability.

```{r}
chains <- lapply(1:3, function(chain) {
  metropolis_hastings(
    log_posterior,
    initial = c(runif(1, -1, 2), runif(1, -2, 1)),
    iterations = 5000,
    proposal_sd = 0.10,
    seed = 430 + chain
  )
})

sapply(chains, function(chain) chain$acceptance_rate)
```

```{r}
#| fig-cap: "Trace plots of the air-yards slope from three Metropolis-Hastings chains started at different points."
#| fig-alt: "Three overlapping noisy lines that converge to the same band after a few hundred iterations."
traces <- lapply(seq_along(chains), function(chain) {
  data.frame(
    chain = factor(chain),
    iteration = seq_len(5000),
    beta1 = chains[[chain]]$draws[, 2]
  )
}) %>%
  bind_rows()

ggplot(traces, aes(iteration, beta1, color = chain)) +
  geom_line(alpha = 0.7, linewidth = 0.3) +
  labs(x = "Iteration", y = "Air-yards slope", color = "Chain") +
  theme_minimal()
```

The chains start far apart and converge to the same region. Discard the first 1,000 iterations as warm-up and summarize the rest.

```{r}
kept <- lapply(chains, function(chain) chain$draws[1001:5000, ])
all_draws <- do.call(rbind, kept)

data.frame(
  parameter = c("beta0", "beta1"),
  mean = colMeans(all_draws),
  sd = apply(all_draws, 2, sd),
  grid_mean = grid_summary$mean
) %>%
  knitr::kable(digits = 3)
```

### Diagnosing a chain

Three quantities describe how useful a chain is.

**Acceptance rate.** For a random-walk proposal in low dimensions, rates around 0.2 to 0.4 are efficient. Much higher rates mean the steps are too small; much lower rates mean most proposals are wasted.

**Effective sample size.** With autocorrelation $\rho_k$ at lag $k$, the $S$ correlated draws carry the information of

$$
S_{\text{eff}}=\frac{S}{1+2\sum_{k\ge1}\rho_k}
$$

independent draws.

**R-hat.** Split each chain in half and compare the variance between the pieces with the variance within them:

$$
\hat R=\sqrt{\frac{\frac{n-1}{n}W+\frac{1}{n}B}{W}},
$$

where $W$ is the mean within-piece variance and $B$ is $n$ times the variance of the piece means. If the chains have mixed, $B$ is small relative to $W$ and $\hat R\approx1$. Values above 1.01 signal that the chains disagree.

```{r}
beta1_chains <- sapply(kept, function(chain) chain[, 2])

data.frame(
  quantity = c("effective sample size (all chains)", "split R-hat"),
  value = c(
    sum(apply(beta1_chains, 2, effective_sample_size)),
    split_rhat(beta1_chains)
  )
) %>%
  knitr::kable(digits = 3)
```

Twelve thousand kept draws yield only about a thousand effective draws. Random-walk proposals explore slowly.

### The proposal scale matters

```{r}
tuning <- lapply(c(0.01, 0.10, 1.00), function(s) {
  chain <- metropolis_hastings(log_posterior, c(0.8, -0.6), 5000, proposal_sd = s, seed = 431)
  data.frame(
    proposal_sd = s,
    acceptance_rate = chain$acceptance_rate,
    effective_sample_size = effective_sample_size(chain$draws[1001:5000, 2])
  )
}) %>%
  bind_rows()

tuning %>% knitr::kable(digits = 3)
```

Tiny steps are almost always accepted but barely move. Huge steps are almost always rejected. The middle scale gives the most effective draws.

## Gibbs sampling

When each parameter's **full conditional** distribution, its distribution given all other parameters and the data, is a known family, we can update parameters one at a time by drawing from those conditionals. No proposal is needed and nothing is rejected.

The normal hierarchical model is the classic case. With passer means $\bar y_j$ over $n_j$ attempts and known within variance $\sigma^2$,

$$
\bar y_j\mid\theta_j\sim N(\theta_j,\sigma^2/n_j),
\qquad
\theta_j\mid\mu,\tau^2\sim N(\mu,\tau^2),
\qquad
p(\mu)\propto1,
\qquad
\tau^2\sim\text{Inv-Gamma}(a_0,b_0),
$$

the full conditionals are

$$
\begin{aligned}
\theta_j\mid\cdot &\sim N\!\left(\frac{n_j\bar y_j/\sigma^2+\mu/\tau^2}{n_j/\sigma^2+1/\tau^2},\ \frac{1}{n_j/\sigma^2+1/\tau^2}\right),\\[4pt]
\mu\mid\cdot &\sim N\!\left(\frac{1}{J}\sum_j\theta_j,\ \frac{\tau^2}{J}\right),\\[4pt]
\tau^2\mid\cdot &\sim\text{Inv-Gamma}\!\left(a_0+\frac{J}{2},\ b_0+\frac12\sum_j(\theta_j-\mu)^2\right).
\end{aligned}
$$

The first line is the normal-normal update from Tutorial 4.4 applied to each passer.

```{r}
qb <- passes %>%
  group_by(passer) %>%
  summarize(n = n(), ybar = mean(qb_epa), .groups = "drop")
sigma2 <- mean(tapply(passes$qb_epa, passes$passer, var), na.rm = TRUE)
J <- nrow(qb)

gibbs <- function(iterations, seed) {
  set.seed(seed)
  a0 <- 1
  b0 <- 0.01
  theta <- qb$ybar
  mu <- mean(theta)
  tau2 <- 0.01
  out <- matrix(NA_real_, iterations, 2 + J)
  for (s in seq_len(iterations)) {
    precision <- qb$n / sigma2 + 1 / tau2
    theta <- rnorm(J, (qb$n * qb$ybar / sigma2 + mu / tau2) / precision, sqrt(1 / precision))
    mu <- rnorm(1, mean(theta), sqrt(tau2 / J))
    tau2 <- 1 / rgamma(1, a0 + J / 2, b0 + sum((theta - mu)^2) / 2)
    out[s, ] <- c(mu, sqrt(tau2), theta)
  }
  colnames(out) <- c("mu", "tau", as.character(qb$passer))
  out
}

gibbs_draws <- gibbs(4000, seed = 431)[1001:4000, ]

intercept_only <- lmer(qb_epa ~ 1 + (1 | passer), data = passes)

data.frame(
  parameter = c("mu", "tau"),
  posterior_mean = colMeans(gibbs_draws[, c("mu", "tau")]),
  posterior_sd = apply(gibbs_draws[, c("mu", "tau")], 2, sd),
  lme4_estimate = c(
    fixef(intercept_only)[["(Intercept)"]],
    as.data.frame(VarCorr(intercept_only))$sdcor[1]
  )
) %>%
  knitr::kable(digits = 4, row.names = FALSE)
```

The posterior means sit near the `lme4` estimates from the intercept-only model of Tutorial 4.2, and now each comes with a posterior standard deviation. The remaining gap comes from the Gibbs sampler's simplification of using per-passer means with a fixed within variance rather than the full attempt-level likelihood.

```{r}
#| fig-cap: "Posterior draws of the between-passer standard deviation from the Gibbs sampler."
#| fig-alt: "Histogram of tau draws, right skewed and centered near 0.09."
ggplot(data.frame(tau = gibbs_draws[, "tau"]), aes(tau)) +
  geom_histogram(bins = 40, fill = "#03a9e6", alpha = 0.8) +
  labs(x = "Between-passer standard deviation", y = "Draws") +
  theme_minimal()
```

Gibbs sampling is exact and simple when the conditionals are available. Its weakness is that strongly correlated parameters are updated one at a time, which makes the chain crawl along a diagonal.

## Hamiltonian Monte Carlo

Hamiltonian Monte Carlo (HMC) treats $\theta$ as the position of a particle and adds an auxiliary momentum $\mathbf r\sim N(\mathbf 0,M)$. The joint density of position and momentum is $\exp\{-H(\theta,\mathbf r)\}$ with

$$
H(\theta,\mathbf r)=-\log p(\theta\mid y)+\tfrac12\mathbf r^{\mathsf T}M^{-1}\mathbf r.
$$

A proposal simulates the particle's motion under this Hamiltonian for $L$ **leapfrog** steps of size $\epsilon$, using the gradient of the log posterior:

$$
\mathbf r\leftarrow\mathbf r+\tfrac{\epsilon}{2}\nabla_\theta\log p(\theta\mid y),
\qquad
\theta\leftarrow\theta+\epsilon M^{-1}\mathbf r,
\qquad
\mathbf r\leftarrow\mathbf r+\tfrac{\epsilon}{2}\nabla_\theta\log p(\theta\mid y).
$$

The endpoint is accepted with probability $\min\{1,\exp(H_{\text{start}}-H_{\text{end}})\}$. Because the trajectory follows the gradient, the proposal can travel far across the posterior while keeping a high acceptance rate, and the number of gradient evaluations grows slowly with dimension.

```{r}
log_posterior_gradient <- function(beta) {
  p <- plogis(as.numeric(X %*% beta))
  as.numeric(t(X) %*% (y - p)) - beta / 2.5^2
}

hmc <- function(iterations, epsilon = 0.02, max_steps = 20, seed = 431) {
  set.seed(seed)
  theta <- c(0.8, -0.6)
  draws <- matrix(NA_real_, iterations, 2)
  accepted <- 0
  for (s in seq_len(iterations)) {
    # Draw the number of leapfrog steps at random so that trajectories of one
    # fixed length cannot fall into a periodic orbit.
    steps <- sample(max_steps, 1)
    r <- rnorm(2)
    theta_new <- theta
    r_new <- r + epsilon / 2 * log_posterior_gradient(theta_new)
    for (l in seq_len(steps)) {
      theta_new <- theta_new + epsilon * r_new
      if (l < steps) r_new <- r_new + epsilon * log_posterior_gradient(theta_new)
    }
    r_new <- r_new + epsilon / 2 * log_posterior_gradient(theta_new)
    h_start <- -log_posterior(theta) + sum(r^2) / 2
    h_end <- -log_posterior(theta_new) + sum(r_new^2) / 2
    if (log(runif(1)) < h_start - h_end) {
      theta <- theta_new
      accepted <- accepted + 1
    }
    draws[s, ] <- theta
  }
  list(draws = draws, acceptance_rate = accepted / iterations)
}

hmc_run <- hmc(2000)

data.frame(
  sampler = c("Metropolis-Hastings, 4000 kept draws", "HMC, 1500 kept draws"),
  acceptance_rate = c(chains[[1]]$acceptance_rate, hmc_run$acceptance_rate),
  effective_sample_size = c(
    effective_sample_size(kept[[1]][, 2]),
    effective_sample_size(hmc_run$draws[501:2000, 2])
  )
) %>%
  knitr::kable(digits = 3)
```

The HMC chain produces far more effective draws per iteration, at the cost of several gradient evaluations per draw. The step size and trajectory length still had to be chosen: with a much longer trajectory the particle circles back toward its start and the effective sample size collapses. Stan's sampler, the No-U-Turn Sampler (NUTS), tunes $\epsilon$, $L$, and $M$ automatically during warm-up, so the user does not set them by hand.

## The multilevel completion model in Stan

Stan is a probabilistic programming language: the model is written as a declaration of data, parameters, and the log posterior, and Stan handles differentiation and sampling. The program below is the crossed completion model from Tutorial 4.3 with priors on every parameter.

```{r}
#| echo: false
#| comment: ""
cat(readLines("R/completion_multilevel.stan"), sep = "\n")
```

Three details are worth noting.

- **Non-centered parameterization.** Each varying intercept is written as $u_j=\tau z_j$ with $z_j\sim N(0,1)$ instead of $u_j\sim N(0,\tau^2)$. The two are the same model, but the second creates a funnel-shaped posterior in $(u_j,\tau)$ that HMC handles poorly when $\tau$ is small.
- **Half-normal priors** on $\tau$ keep the standard deviations away from implausibly large values while allowing them to be near zero.
- `bernoulli_logit` combines the sigmoid and the Bernoulli likelihood in one numerically stable call.

The fit was run with `cmdstanr` (4 chains, 1,000 warm-up and 1,000 sampling iterations each) and stored in `data/bayes_results.rds`.

```{r}
#| eval: false
library(cmdstanr)
model <- cmdstan_model("R/completion_multilevel.stan")
fit <- model$sample(
  data = stan_data,
  seed = 431,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000
)
```

### Hyperparameters and diagnostics

```{r}
results$stan$hyper_summary %>%
  select(variable, mean, sd, q5, q95, rhat, ess_bulk) %>%
  knitr::kable(digits = 3)
```

```{r}
results$stan$diagnostics %>% knitr::kable()
```

Every R-hat is close to 1, the bulk effective sample sizes are in the hundreds or thousands, and there are no divergent transitions. A divergent transition is HMC's warning that the trajectory left the region where the numerical integrator is accurate, which usually signals a funnel or another awkward geometry.

```{r}
#| fig-cap: "Trace of the passer standard deviation across the four Stan chains."
#| fig-alt: "Four overlapping traces mixing well around 0.1."
tau_chains <- results$stan$tau_qb_chains
data.frame(
  iteration = rep(seq_len(nrow(tau_chains)), 4),
  chain = factor(rep(1:4, each = nrow(tau_chains))),
  tau_qb = as.numeric(tau_chains)
) %>%
  ggplot(aes(iteration, tau_qb, color = chain)) +
  geom_line(alpha = 0.7, linewidth = 0.3) +
  labs(x = "Sampling iteration", y = "Passer standard deviation", color = "Chain") +
  theme_minimal()
```

### Bayesian effects against lme4 estimates

```{r}
qb_posterior <- data.frame(
  passer = colnames(results$stan$u_qb_draws),
  posterior_mean = colMeans(results$stan$u_qb_draws),
  posterior_sd = apply(results$stan$u_qb_draws, 2, sd),
  lower_90 = apply(results$stan$u_qb_draws, 2, quantile, 0.05),
  upper_90 = apply(results$stan$u_qb_draws, 2, quantile, 0.95)
) %>%
  left_join(
    results$lme4$completion_qb_effects %>%
      tibble::rownames_to_column("passer") %>%
      rename(lme4_effect = `(Intercept)`),
    by = "passer"
  ) %>%
  left_join(results$qb_attempts %>% mutate(passer = as.character(passer)), by = "passer")

qb_posterior %>%
  filter(attempts >= 300) %>%
  arrange(desc(posterior_mean)) %>%
  select(passer, attempts, posterior_mean, posterior_sd, lower_90, upper_90, lme4_effect) %>%
  slice_head(n = 8) %>%
  knitr::kable(digits = 3)
```

```{r}
#| fig-cap: "Posterior means from Stan against lme4 estimates for every passer. The two agree closely; the Bayesian version adds a full posterior for each."
#| fig-alt: "Scatter plot along the diagonal."
ggplot(qb_posterior, aes(lme4_effect, posterior_mean)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(aes(size = attempts), color = "#03a9e6", alpha = 0.7) +
  labs(x = "lme4 estimate", y = "Stan posterior mean", size = "Attempts") +
  theme_minimal()
```

Because the data are plentiful, empirical Bayes and full Bayes agree on the point estimates. The value of the posterior draws is in what they let us compute next.

### Probabilities that the data cannot express as point estimates

```{r}
draws <- results$stan$u_qb_draws
pair <- c("Brock Purdy", "Patrick Mahomes")
mean(draws[, pair[[1]]] > draws[, pair[[2]]])
```

That number is $P(u_{\text{Purdy}}>u_{\text{Mahomes}}\mid y)$, the posterior probability that one passer's completion ability exceeds the other's after adjusting for throw difficulty, receivers, and opponents. It is a direct answer to a question that a point estimate and a standard error can only approximate.

```{r}
volume <- results$qb_attempts %>% filter(attempts >= 300) %>% pull(passer) %>% as.character()
ranks <- rank_probabilities(draws[, volume])

data.frame(
  passer = colnames(ranks),
  probability_top_5 = colMeans(ranks <= 5),
  expected_rank = colMeans(ranks)
) %>%
  arrange(desc(probability_top_5)) %>%
  slice_head(n = 8) %>%
  knitr::kable(digits = 2, row.names = FALSE)
```

No passer is a lock for the top five. Rankings of adjusted player effects carry substantial uncertainty even in a full season, and a table of rank probabilities communicates that honestly.

## Practice

1. Why can the Metropolis acceptance ratio be computed without the normalizing constant?
2. A chain with 10,000 draws has lag-1 autocorrelation 0.9 and negligible autocorrelation beyond that. Approximately what is its effective sample size?
3. Why does the non-centered parameterization help when $\tau$ is small?
4. State one question the posterior draws answer that the `lme4` output does not.

::: {.callout-tip collapse="true" title="Show answers"}
1. The ratio $p(\theta^{*}\mid y)/p(\theta\mid y)$ equals $p(y\mid\theta^{*})p(\theta^{*})/[p(y\mid\theta)p(\theta)]$; the constant $p(y)$ appears in both numerator and denominator and cancels.
2. $10{,}000/(1+2\times0.9)\approx3{,}570$.
3. When $\tau$ is small, $u_j$ must be small too, so the joint posterior of $(u_j,\tau)$ is a narrow funnel that a fixed step size cannot traverse. Writing $u_j=\tau z_j$ makes $z_j$ approximately standard normal regardless of $\tau$.
4. For example, the probability that one passer is better than another, or the probability that a passer is among the top five.
:::

## Key takeaways

- Posterior summaries are integrals; grids work in one or two dimensions, the Laplace approximation works for near-normal posteriors, and MCMC works in general.
- Metropolis-Hastings needs only the unnormalized posterior; its efficiency depends on the proposal scale.
- Gibbs sampling draws from full conditionals and fits the normal hierarchical model in a few lines.
- HMC uses gradients to propose distant states with high acceptance, and Stan automates it.
- Always check R-hat, effective sample size, and divergences before reading a posterior.
- In this example the Bayesian posterior means are close to the `lme4` conditional modes. The draws support probabilities of comparisons and rankings.
