---
title: "6.1 Unsupervised Learning: Goals, Features, and Distances"
subtitle: "What a clustering can claim, and the decisions that determine it"
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
source("R/cluster_helpers.R")

nba <- load_nba()
profiles <- nba$profiles
```

## Learning objectives

After this tutorial, you should be able to:

- distinguish unsupervised from supervised learning and state what a clustering can and cannot claim;
- separate the two purposes of clustering players, archetypes and comparables, and explain why they call for different methods;
- construct a dissimilarity deliberately through representation, transformation, standardization, weighting, and aggregation;
- compute and compare Euclidean, Manhattan, correlation, Mahalanobis, and Gower distances; and
- use principal components to inspect and reduce a feature matrix before clustering.

## Learning without a target

Lectures 2 through 5 were supervised: every model had an outcome, and a held-out outcome told us whether the model was any good. Clustering has no outcome. We choose what "similar" means, an algorithm groups similar observations, and the only judge of the result is whether it is stable, interpretable, and useful for the question that motivated it.

That changes what a clustering can claim. "These players form five groups" is a statement about our distance, our features, and our algorithm as much as about the players. Three consequences follow.

- The **features** and the **distance** are modeling decisions and should be designed for the question, not left at software defaults.
- The **number of clusters** is not a fact to be discovered; it is a resolution to be chosen, and different resolutions answer different questions.
- **Validation** means checking that the structure is real (not an artifact of the algorithm) and stable (not an artifact of the sample), which is the subject of Tutorial 6.6.

## Two purposes: archetypes and comparables

Akhanli and Hennig (2023) distinguish two uses of player clustering that need different tools.

| Purpose | Number of clusters | Cluster size | Typical use | What matters |
|---|---|---|---|---|
| **Archetypes** | few (4 to 8) | large | describing roster construction, lineup balance, how the game is played | well-separated, interpretable groups |
| **Comparables** | many (20 to 50) | small | scouting, contract valuation, projecting a player from similar players | tight within-cluster similarity |

A method tuned for archetypes will produce coarse groups that lump distinct roles together; a method tuned for comparables will split a real archetype into many pieces. Decide which you want before running anything.

## The data

```{r}
profiles %>%
  select(player, position, team, games, minutes, pts_36, fg3a_36, ast_36, dreb_36, blk_36, three_rate, efg) %>%
  slice_head(n = 8) %>%
  knitr::kable(digits = 2)
```

Every player played at least 500 minutes in 2023-24. The listed position is kept as an external reference for later tutorials and is never used as an input.

```{r}
data.frame(feature = profile_features, label = feature_labels[profile_features]) %>%
  knitr::kable(row.names = FALSE)
```

## Five decisions in a dissimilarity

Akhanli and Hennig describe a dissimilarity as the result of five explicit steps. Each one changes the clustering.

### 1. Representation

Totals confound role with playing time: a starter and a backup with identical per-minute production have very different season totals. Per-36-minute rates remove playing time. Shares such as three-point attempts divided by all attempts remove volume as well, leaving shot *selection*. Which representation is right depends on the question: for archetypes of play style, rates and shares; for a player's total contribution, totals.

```{r}
profiles %>%
  select(player, minutes, fg3a, fg3a_36, three_rate) %>%
  arrange(desc(fg3a)) %>%
  slice_head(n = 5) %>%
  knitr::kable(digits = 2)
```

### 2. Transformation

A difference of one block per 36 minutes means something different near zero than it does at three. Skewed variables let a few extreme players dominate distances.

```{r}
#| fig-cap: "Blocks per 36 minutes before and after a square-root transformation."
#| fig-alt: "Two histograms; the raw distribution has a long right tail, the transformed one is closer to symmetric."
bind_rows(
  data.frame(value = profiles$blk_36, scale = "blocks per 36"),
  data.frame(value = sqrt(profiles$blk_36), scale = "square root of blocks per 36")
) %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 30, fill = "#03a9e6", alpha = 0.8) +
  facet_wrap(~scale, scales = "free_x") +
  labs(x = NULL, y = "Players") +
  theme_minimal()
```

The transformation is a statement about which differences matter. Under the square root, the gap between 0 and 1 block counts more than the gap between 3 and 4, which is usually the intent.

### 3. Standardization

Points per 36 minutes ranges over about 30 units; effective field-goal percentage ranges over about 0.3. Without rescaling, the points variable alone would decide the distance. The usual fix is the z-score,

$$
z_{ik}=\frac{x_{ik}-\bar x_k}{s_k},
$$

which gives every variable unit variance. Alternatives are range scaling, $(x-\min)/(\max-\min)$, and robust scaling by the median and median absolute deviation, which resists outliers.

```{r}
X <- profile_matrix(profiles)
round(apply(X, 2, sd)[1:6], 3)
```

### 4. Weighting

Standardization makes every variable count equally, which is itself a choice. If the question is about offensive role, shot selection variables deserve more weight than steals. A weighted Euclidean distance is

$$
d_w(\mathbf x_i,\mathbf x_j)=\sqrt{\sum_{k=1}^{p}w_k\left(z_{ik}-z_{jk}\right)^2},
$$

and multiplying column $k$ by $\sqrt{w_k}$ before an unweighted algorithm has the same effect. Weights also handle redundancy: including points, field-goal attempts, and free-throw attempts triple-counts scoring volume unless their weights are reduced.

### 5. Aggregation

The Minkowski family combines per-variable differences,

$$
d_q(\mathbf x_i,\mathbf x_j)=\left(\sum_{k=1}^{p}w_k\left|z_{ik}-z_{jk}\right|^q\right)^{1/q},
$$

with $q=2$ (Euclidean) and $q=1$ (Manhattan) the common cases. Euclidean distance emphasizes large differences on single variables; Manhattan distance treats a difference of 2 on one variable the same as differences of 1 on two variables.

## Distances that measure different things

### Correlation distance

Two players with the same *pattern* of strengths but different overall levels are close under the correlation distance

$$
d_{\text{cor}}(\mathbf x_i,\mathbf x_j)=1-r(\mathbf x_i,\mathbf x_j),
$$

where $r$ is the Pearson correlation between their standardized feature vectors. A high-usage scorer and a low-usage scorer with the same shape of profile are neighbors here and far apart under Euclidean distance.

### Mahalanobis distance

Correlated features double-count. Points, field-goal attempts, and free-throw attempts move together, so Euclidean distance on all three is roughly three times the distance on one. The Mahalanobis distance

$$
d_M(\mathbf x_i,\mathbf x_j)=\sqrt{(\mathbf x_i-\mathbf x_j)^{\mathsf T}S^{-1}(\mathbf x_i-\mathbf x_j)}
$$

divides by the covariance matrix $S$, which removes the double-counting. It is Euclidean distance after decorrelating the features, and it is the natural distance inside a Gaussian mixture model (Tutorial 6.4).

### Gower distance for mixed data

When some variables are categorical, such as handedness or position, Gower's coefficient averages per-variable dissimilarities that are each scaled to $[0,1]$: absolute differences divided by the range for numeric variables and a 0/1 mismatch for categorical ones. `cluster::daisy()` computes it; Tutorial 6.5 uses it.

### The same player, three sets of neighbors

```{r}
nearest <- function(D, name, k = 5) {
  D <- as.matrix(D)
  ordered <- sort(D[name, ])
  names(ordered)[2:(k + 1)]
}

euclidean <- dist(X)
manhattan <- dist(X, method = "manhattan")
correlation <- as.dist(1 - cor(t(X)))
mahalanobis_matrix <- {
  S_inverse <- solve(cov(X))
  as.dist(sqrt(outer(seq_len(nrow(X)), seq_len(nrow(X)), Vectorize(function(i, j) {
    d <- X[i, ] - X[j, ]
    as.numeric(t(d) %*% S_inverse %*% d)
  }))))
}
attr(mahalanobis_matrix, "Labels") <- rownames(X)

target <- "Nikola Jokic"
data.frame(
  euclidean = nearest(euclidean, target),
  manhattan = nearest(manhattan, target),
  correlation = nearest(correlation, target),
  mahalanobis = nearest(mahalanobis_matrix, target)
) %>%
  knitr::kable()
```

```{r}
target <- "Stephen Curry"
data.frame(
  euclidean = nearest(euclidean, target),
  manhattan = nearest(manhattan, target),
  correlation = nearest(correlation, target),
  mahalanobis = nearest(mahalanobis_matrix, target)
) %>%
  knitr::kable()
```

The lists overlap but are not the same. The neighbors are a property of the distance, and the distance should be chosen to match what "comparable" means for the question.

::: {.callout-important}
## Defaults are not decisions

`dist()` returns Euclidean distance on whatever columns it is given. That is a default, not a recommendation. Every step above, from per-minute rates to the choice of $q$, changes which players are neighbors.
:::

## The curse of dimensionality

In high dimensions, distances lose contrast: the nearest and farthest points from any observation become nearly equally far. The following simulation draws points uniformly in a cube and records the ratio of the farthest to the nearest neighbor of one point.

```{r}
set.seed(431)
sapply(c(2, 5, 10, 14, 50, 200), function(p) {
  points <- matrix(runif(300 * p), ncol = p)
  d <- sqrt(colSums((t(points[-1, ]) - points[1, ])^2))
  c(dimensions = p, farthest_over_nearest = round(max(d) / min(d), 2))
}) %>%
  t() %>%
  knitr::kable()
```

With 14 features the problem is mild but present. Redundant features make it worse without adding information, which is one argument for weighting, for Mahalanobis distance, and for dimension reduction.

## Principal components

Principal component analysis (PCA) finds the directions of largest variance in the standardized feature matrix. With $Z$ the $n\times p$ standardized matrix and $S=Z^{\mathsf T}Z/(n-1)$ its correlation matrix, the eigendecomposition $S=V\Lambda V^{\mathsf T}$ gives loadings $V$ and variances $\Lambda=\operatorname{diag}(\lambda_1\ge\cdots\ge\lambda_p)$. The scores $T=ZV$ are uncorrelated new variables, and the share of variance carried by the first $m$ components is $\sum_{k\le m}\lambda_k/\sum_k\lambda_k$.

```{r}
pca <- prcomp(X, center = FALSE, scale. = FALSE)
variance <- pca$sdev^2 / sum(pca$sdev^2)

data.frame(
  component = seq_along(variance),
  variance_share = variance,
  cumulative = cumsum(variance)
) %>%
  slice_head(n = 8) %>%
  knitr::kable(digits = 3)
```

```{r}
loadings <- pca$rotation[, 1:3]
loadings %>%
  as.data.frame() %>%
  tibble::rownames_to_column("feature") %>%
  mutate(feature = feature_labels[feature]) %>%
  knitr::kable(digits = 2)
```

The first component contrasts interior play (rebounds, blocks, free-throw rate) with perimeter play (three-point rate, assists). The second component is scoring volume. Together they describe a large share of the variation among players.

```{r}
#| fig-cap: "Players on the first two principal components, colored by listed position group. Positions are shown for orientation only; they were not used in the analysis."
#| fig-alt: "Scatter plot with guards on one side, centers on the other, and forwards in between, with considerable overlap."
scores <- data.frame(pca$x[, 1:2], player = profiles$player, position_group = profiles$position_group, minutes = profiles$minutes)

ggplot(scores, aes(PC1, PC2, color = position_group)) +
  geom_point(alpha = 0.7, size = 2) +
  ggrepel::geom_text_repel(
    data = scores %>% arrange(desc(minutes)) %>% slice_head(n = 12),
    aes(label = player), size = 2.6, color = "black", max.overlaps = 20
  ) +
  labs(x = paste0("PC1 (", round(100 * variance[1]), "% of variance)"),
       y = paste0("PC2 (", round(100 * variance[2]), "% of variance)"),
       color = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Two things stand out. Position groups occupy different regions but overlap heavily, so any clustering will disagree with the labels for many players. And there are no empty gaps between groups: the players form a continuum with denser and sparser regions, not islands. Clustering will impose boundaries on that continuum, which is why validation matters.

### PCA as preprocessing

Clustering on the first few components has three effects: it removes noise in the small components, it decorrelates the features (so Euclidean distance on scores is Mahalanobis-like on the original variables when all components are kept), and it makes the result plottable. It also has a cost: PCA maximizes variance, not cluster separation, and a direction that separates two rare roles can be discarded as low-variance. Keeping enough components to explain about 80 to 90% of variance is a common compromise, and Tutorial 6.6 checks whether the clusters change.

## Practice

1. Why do per-36 rates rather than totals suit a study of play style, and when would totals be preferable?
2. Two players have profiles $(3, 1, 1)$ and $(1, 3, 1)$ on three standardized variables. Compute their Euclidean, Manhattan, and correlation distances.
3. Explain why Mahalanobis distance can be preferable to Euclidean when several features measure scoring volume.
4. What is lost when clustering is done on the first two principal components only?

::: {.callout-tip collapse="true" title="Show answers"}
1. Rates remove playing time, which is a role decision made by the coach rather than a characteristic of the player's style; totals are preferable when the question concerns total contribution or availability.
2. Euclidean $\sqrt{4+4+0}=2.83$; Manhattan $2+2+0=4$; the vectors have correlation $-0.5$ so the correlation distance is $1.5$.
3. Correlated features count the same information several times in a Euclidean distance; the covariance inverse removes that redundancy so each independent dimension of variation counts once.
4. Directions of small variance that may nonetheless separate real groups, and any structure orthogonal to the two retained components.
:::

## Key takeaways

- Clustering has no outcome to check against; the features and distance are modeling decisions that define the answer.
- Decide first whether the goal is a few archetypes or many comparables.
- Build a dissimilarity in five deliberate steps: representation, transformation, standardization, weighting, and aggregation.
- Euclidean, Manhattan, correlation, Mahalanobis, and Gower distances answer different questions and produce different neighbors.
- Principal components reveal that players form a continuum; clustering draws boundaries on it and must be validated.
