---
title: "6.2 k-Means and k-Medoids"
subtitle: "The objective, Lloyd's algorithm, seeding, choosing k, and player archetypes"
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
library(cluster)
source("R/cluster_helpers.R")

nba <- load_nba()
profiles <- nba$profiles
X <- profile_matrix(profiles)
```

## Learning objectives

After this tutorial, you should be able to:

- state the k-means objective and show why each step of Lloyd's algorithm cannot increase it;
- implement Lloyd's algorithm and k-means++ seeding and explain why multiple starts are needed;
- choose the number of clusters with the elbow, silhouette, and gap statistic and say what each one measures;
- interpret k-means clusters as player archetypes and identify the players that fit poorly;
- use k-medoids (PAM) with an arbitrary dissimilarity; and
- list the assumptions built into k-means.

## The objective

Given $n$ observations $\mathbf x_1,\dots,\mathbf x_n\in\mathbb R^p$, k-means seeks a partition $C_1,\dots,C_k$ and centers $\boldsymbol\mu_1,\dots,\boldsymbol\mu_k$ that minimize the within-cluster sum of squares,

$$
W(C,\boldsymbol\mu)=\sum_{j=1}^{k}\sum_{i\in C_j}\lVert\mathbf x_i-\boldsymbol\mu_j\rVert^2.
$$

For a fixed partition the best center is the cluster mean, and $W$ then equals half the average squared distance between members of the same cluster:

$$
\sum_{i\in C_j}\lVert\mathbf x_i-\bar{\mathbf x}_j\rVert^2
=\frac{1}{2|C_j|}\sum_{i\in C_j}\sum_{i'\in C_j}\lVert\mathbf x_i-\mathbf x_{i'}\rVert^2.
$$

So k-means is a statement about pairwise Euclidean distances within clusters. Minimizing $W$ exactly is NP-hard; every practical algorithm finds a local optimum.

## Lloyd's algorithm

Lloyd's algorithm alternates two steps from an initial set of centers.

1. **Assignment.** Give each observation to its nearest center: $C_j=\{i:\ j=\arg\min_l\lVert\mathbf x_i-\boldsymbol\mu_l\rVert\}$.
2. **Update.** Replace each center by the mean of its members: $\boldsymbol\mu_j=\frac{1}{|C_j|}\sum_{i\in C_j}\mathbf x_i$.

Each step minimizes $W$ over one set of variables with the other held fixed, so $W$ never increases, and because there are finitely many partitions the algorithm terminates. Where it terminates depends on where it started.

```{r}
set.seed(431)
scratch <- lloyd_kmeans(X, k = 5, seed = 431)

data.frame(iteration = seq_along(scratch$wcss), within_ss = scratch$wcss) %>%
  knitr::kable(digits = 1)
```

```{r}
package_fit <- kmeans(X, centers = 5, nstart = 25, iter.max = 100)
c(scratch_within_ss = tail(scratch$wcss, 1), package_within_ss = package_fit$tot.withinss)
```

The from-scratch and package solutions reach similar objective values. With many restarts, `kmeans()` usually finds a slightly better local optimum.

### Local optima and multiple starts

```{r}
#| fig-cap: "Final within-cluster sum of squares over 60 random starts, with random seeding and with k-means++ seeding."
#| fig-alt: "Two overlapping histograms; the k-means++ distribution is concentrated at lower values."
starts <- bind_rows(
  lapply(1:60, function(s) data.frame(seeding = "random", within_ss = tail(lloyd_kmeans(X, 5, seed = s, init = "random")$wcss, 1))),
  lapply(1:60, function(s) data.frame(seeding = "k-means++", within_ss = tail(lloyd_kmeans(X, 5, seed = s, init = "plusplus")$wcss, 1)))
)

ggplot(starts, aes(within_ss, fill = seeding)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  labs(x = "Within-cluster sum of squares at convergence", y = "Starts", fill = NULL) +
  theme_minimal()
```

Different starts end at different partitions with different objective values. The standard practice is to run many starts and keep the best, which is what the `nstart` argument does.

### k-means++ seeding

Random seeding often places two initial centers in the same dense region. k-means++ (Arthur and Vassilvitskii, 2007) chooses the first center uniformly and each later center with probability proportional to its squared distance from the nearest center already chosen,

$$
P(\text{choose }\mathbf x_i)=\frac{D(\mathbf x_i)^2}{\sum_{i'}D(\mathbf x_{i'})^2},
$$

which spreads the seeds out. The expected objective after seeding alone is within a factor $O(\log k)$ of the optimum, and Lloyd iterations from those seeds converge faster and to better solutions on average.

## Choosing k

Three common criteria answer slightly different questions.

### The elbow

Plot $W_k$, the optimized objective for each $k$. It always decreases; the "elbow" is where additional clusters stop buying much.

```{r}
k_values <- 1:10
within <- sapply(k_values, function(k) kmeans(X, k, nstart = 25, iter.max = 100)$tot.withinss)
```

### The silhouette

For observation $i$ in cluster $A$, let $a_i$ be its mean distance to the other members of $A$ and $b_i$ the smallest mean distance to the members of any other cluster. The silhouette width is

$$
s_i=\frac{b_i-a_i}{\max(a_i,b_i)}\in[-1,1].
$$

Values near 1 mean the observation sits well inside its cluster, near 0 that it lies between two clusters, and below 0 that it is probably in the wrong cluster. The average silhouette over all observations scores a clustering, and the $k$ with the largest average is a common choice.

```{r}
D <- dist(X)
silhouette_means <- sapply(2:10, function(k) {
  fit <- kmeans(X, k, nstart = 25, iter.max = 100)
  mean(silhouette(fit$cluster, D)[, "sil_width"])
})
```

### The gap statistic

The gap statistic (Tibshirani, Walther, and Hastie, 2001) compares $\log W_k$ with its expected value under a reference distribution with no cluster structure, typically uniform over the range of each feature:

$$
\operatorname{Gap}(k)=E^{*}\!\left[\log W_k^{*}\right]-\log W_k,
$$

where the expectation is estimated from $B$ reference data sets. The recommended rule chooses the smallest $k$ such that $\operatorname{Gap}(k)\ge\operatorname{Gap}(k+1)-s_{k+1}$, where $s_{k+1}$ is the standard error of the reference simulations.

```{r}
set.seed(431)
gap <- clusGap(X, FUN = function(x, k) kmeans(x, k, nstart = 10, iter.max = 100), K.max = 10, B = 50)
gap_table <- as.data.frame(gap$Tab)
gap_k <- maxSE(gap_table$gap, gap_table$SE.sim, method = "firstSEmax")
```

```{r}
#| fig-cap: "Three criteria for the number of clusters on the standardized player profiles."
#| fig-alt: "Three panels: a decreasing curve of within-cluster sum of squares, a silhouette curve peaking at small k, and a gap curve."
criteria <- bind_rows(
  data.frame(k = k_values, value = within, criterion = "within-cluster SS"),
  data.frame(k = 2:10, value = silhouette_means, criterion = "average silhouette"),
  data.frame(k = 1:10, value = gap_table$gap, criterion = "gap statistic")
)

ggplot(criteria, aes(k, value)) +
  geom_line(color = "#03a9e6") +
  geom_point(color = "#03a9e6") +
  facet_wrap(~criterion, scales = "free_y") +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Number of clusters", y = NULL) +
  theme_minimal()
```

```{r}
data.frame(
  criterion = c("average silhouette", "gap statistic (first SE max)"),
  chosen_k = c(which.max(silhouette_means) + 1, gap_k)
) %>%
  knitr::kable()
```

The criteria disagree, and that is the normal outcome on data without well-separated islands. The silhouette usually favors a small $k$ that splits the continuum at its widest gap; the gap statistic and elbow are more generous. When the purpose is archetypes, a $k$ between 4 and 8 that yields interpretable, reasonably stable groups is the practical target, and Tutorial 6.6 measures the stability.

## Player archetypes with k = 6

```{r}
set.seed(431)
k <- 6
fit <- kmeans(X, k, nstart = 50, iter.max = 100)
profiles$cluster <- factor(fit$cluster)

profiles %>%
  count(cluster, name = "players") %>%
  knitr::kable()
```

### Cluster profiles

The centers are the mean standardized feature values of each cluster. A heat map of the centers is the fastest way to name the clusters.

```{r}
#| fig-cap: "Cluster centers in standard-deviation units for each feature."
#| fig-alt: "Heat map with clusters as columns and features as rows, colored from blue (below average) to red (above average)."
centers <- as.data.frame(fit$centers) %>%
  mutate(cluster = factor(seq_len(k))) %>%
  tidyr::pivot_longer(-cluster, names_to = "feature", values_to = "z") %>%
  mutate(feature = factor(feature_labels[feature], levels = rev(feature_labels[profile_features])))

ggplot(centers, aes(cluster, feature, fill = z)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.1f", z)), size = 2.8) +
  scale_fill_gradient2(low = "#4575b4", mid = "white", high = "#d73027", midpoint = 0) +
  labs(x = "Cluster", y = NULL, fill = "z-score") +
  theme_minimal()
```

```{r}
profiles %>%
  group_by(cluster) %>%
  arrange(desc(minutes)) %>%
  summarize(examples = paste(head(player, 4), collapse = ", "), .groups = "drop") %>%
  knitr::kable()
```

Reading the columns of the heat map gives names such as "high-usage creators", "rim-protecting bigs", "spot-up wings", and "low-usage ball handlers". Naming is interpretation, and a good name should survive a look at the example players.

### Agreement with listed positions

```{r}
table(cluster = profiles$cluster, position = profiles$position_group) %>%
  knitr::kable()
```

The clusters track positions loosely. Centers concentrate in one or two clusters, while forwards and guards are spread across several. That is expected: the profiles describe *what players do*, and modern positions describe that only partly.

### Players that fit poorly

```{r}
sil <- silhouette(fit$cluster, D)
profiles$silhouette <- sil[, "sil_width"]

profiles %>%
  arrange(silhouette) %>%
  select(player, position, cluster, silhouette, pts_36, ast_36, dreb_36, three_rate) %>%
  slice_head(n = 8) %>%
  knitr::kable(digits = 2)
```

Negative or near-zero silhouettes mark players between archetypes. These are often the most interesting players in the league: their profiles do not resemble a type, which is precisely why a hard partition struggles with them. Tutorial 6.4 replaces hard assignments with probabilities.

## k-medoids

k-means requires Euclidean distance because the mean minimizes squared Euclidean deviations. **k-medoids** replaces the mean with a **medoid**, the member of the cluster with the smallest total dissimilarity to the others, and minimizes

$$
\sum_{j=1}^{k}\sum_{i\in C_j}d(\mathbf x_i,\mathbf m_j),
$$

where $\mathbf m_j$ is an actual observation. Because only dissimilarities between observations are needed, any distance from Tutorial 6.1 can be used, and because the objective sums distances rather than squared distances, outliers pull less. The standard algorithm is Partitioning Around Medoids (PAM), which swaps medoids with non-medoids while the objective improves.

```{r}
pam_fit <- pam(dist(X, method = "manhattan"), k = k)

data.frame(
  cluster = seq_len(k),
  medoid = rownames(X)[pam_fit$id.med],
  size = as.numeric(table(pam_fit$clustering))
) %>%
  knitr::kable()
```

The medoids are real players, which makes each cluster easy to describe: "players like this one."

```{r}
data.frame(
  comparison = "k-means (Euclidean) versus PAM (Manhattan)",
  adjusted_rand_index = adjusted_rand_index(fit$cluster, pam_fit$clustering)
) %>%
  knitr::kable(digits = 2)
```

The adjusted Rand index, defined in Tutorial 6.6, is 1 for identical partitions and about 0 for unrelated ones. The two methods agree on much of the structure and differ at the boundaries, which is where the choice of distance matters.

## What k-means assumes

- **Clusters are spherical and similar in size** in the standardized space, because every observation goes to the nearest center by Euclidean distance.
- **Every observation belongs to exactly one cluster**, with no notion of uncertainty or of outliers.
- **$k$ is known.** The algorithm has no internal reason to prefer one $k$ over another.
- **The scale of the features is meaningful**, so standardization and weighting decide the result.

Each of the next three tutorials relaxes one of these: hierarchical clustering removes the fixed $k$, mixture models add elongated clusters and soft membership, and density-based methods add outliers and arbitrary shapes.

## Practice

1. Show that for a fixed cluster the mean minimizes $\sum_{i\in C}\lVert\mathbf x_i-\boldsymbol\mu\rVert^2$ over $\boldsymbol\mu$.
2. Why does the within-cluster sum of squares always decrease with $k$, and why does that make it unusable on its own for choosing $k$?
3. An observation has $a_i=2$ and $b_i=3$. What is its silhouette, and how would you describe its position?
4. Give one situation in which PAM is preferable to k-means.

::: {.callout-tip collapse="true" title="Show answers"}
1. The gradient with respect to $\boldsymbol\mu$ is $-2\sum_{i\in C}(\mathbf x_i-\boldsymbol\mu)$, which is zero only at the cluster mean, and the objective is convex in $\boldsymbol\mu$.
2. With $k+1$ clusters one can always reproduce the best $k$-cluster solution and split one cluster, which cannot increase the objective; a criterion that always improves cannot select a value.
3. $s_i=(3-2)/3=0.33$: inside its cluster but not far from the next one.
4. When the dissimilarity is not Euclidean (Manhattan, Gower, correlation), when outliers are present, or when an interpretable representative observation is wanted for each cluster.
:::

## Key takeaways

- k-means minimizes within-cluster squared Euclidean distance; Lloyd's algorithm finds a local optimum, so use many starts and k-means++ seeding.
- The elbow, silhouette, and gap statistic measure different things and often disagree; choose $k$ for the purpose.
- Cluster centers in z-score units are the primary tool for naming archetypes; silhouettes identify the players that fit poorly.
- k-medoids works with any dissimilarity and returns real players as representatives.
