---
title: "6.7 Guided Lab: Fuzzy Roles and Uncertain Shot Profiles"
subtitle: "Compositional distances, fuzzy c-means, Bayesian measurement uncertainty, and consensus"
---

## Learning objectives

You will construct player shot-selection profiles, explain why shares require special treatment, fit fuzzy c-means, distinguish fuzzy membership from a probability, and propagate uncertainty in observed shot counts into a consensus summary. This extends the density, spectral, mixed-data, and NMF methods in Tutorial 6.5.

## Sports question and unit of analysis

Which players have similar shot-selection styles, and how confident should we be when a player has relatively few attempts? A player-season is the unit of analysis. A shot is a repeated observation within that unit. Shot selection differs from shooting accuracy: this lab uses attempts from locations, regardless of whether they went in.

We use three broad zones to make the geometry visible: rim attempts within 4 feet, other two-point attempts, and three-point attempts. These are teaching categories, not an official taxonomy. The stored `three` indicator takes precedence over the radius rule. Select players with at least 200 recorded attempts to reduce severe measurement noise. This cutoff defines the population described by the results.

```{r}
library(dplyr)
library(ggplot2)
source("R/cluster_helpers.R")
nba <- load_nba()
zone_data <- nba$shots %>% mutate(zone = case_when(
  three == 1 ~ "three", distance <= 4 ~ "rim", TRUE ~ "other_two"))
zone_data$zone <- factor(zone_data$zone, levels = c("rim", "other_two", "three"))
counts <- unclass(xtabs(~ athlete_id + zone, data = zone_data))
counts <- counts[rowSums(counts) >= 200, , drop = FALSE]
player_key <- zone_data %>% distinct(athlete_id, shooter) %>%
  distinct(athlete_id, .keep_all = TRUE)
player_names <- player_key$shooter[match(rownames(counts), as.character(player_key$athlete_id))]
stopifnot(!anyNA(player_names), all(rowSums(counts) >= 200))
head(data.frame(player = player_names, attempts = rowSums(counts), counts))
```

## A composition lives on a simplex

Let $c_{iz}$ be the shot count for player $i$ in zone $z$, $n_i=\sum_z c_{iz}$, and $p_{iz}$ the underlying shot-selection probability. The vector $\mathbf p_i$ has $D=3$ components and satisfies $p_{iz}\ge0$ and $\sum_zp_{iz}=1$. Increasing one share necessarily decreases at least one other share. Independent unconstrained normal modeling of all shares ignores that constraint.

The **centered log-ratio** (CLR) transform is

$$
\operatorname{clr}(\mathbf p_i)_z=\log p_{iz}-\frac1D\sum_{v=1}^{D}\log p_{iv}
=\log\frac{p_{iz}}{g(\mathbf p_i)},
$$

where $g(\mathbf p_i)$ is the geometric mean of the shares. Euclidean distance between CLR vectors is the Aitchison distance. It compares relative allocations. A difference between 1% and 2% can matter differently from a difference between 40% and 41% because the ratios differ.

CLR coordinates sum to zero, so their covariance matrix is singular. k-means can operate on their distances, but an unrestricted full-covariance GMM should use $D-1$ orthonormal log-ratio coordinates instead. PCA on CLR coordinates supplies a convenient orthogonal basis for the observed contrast subspace in this demonstration.

Zeros prevent taking logarithms. We use a Dirichlet prior with $\alpha_z=0.5$ per zone, giving the posterior mean

$$
\widehat p_{iz}=\frac{c_{iz}+\alpha_z}{n_i+\sum_v\alpha_v}.
$$

This is an explicit smoothing assumption, not evidence that a player actually attempted every shot type. A structural zero, such as a physically unavailable action, requires different treatment from a zone that happened to have zero observed attempts.

```{r}
alpha <- rep(0.5, ncol(counts))
posterior_shape <- sweep(counts, 2, alpha, "+")
P <- posterior_shape / rowSums(posterior_shape)
clr <- function(P) log(P) - rowMeans(log(P))
Z <- clr(P)
stopifnot(max(abs(rowSums(Z))) < 1e-10)
projection <- prcomp(Z, center = TRUE, scale. = FALSE)
coordinates <- projection$x[, 1:2]
set.seed(631)
hard <- kmeans(Z, centers = 4, nstart = 50)
```

We do not standardize CLR columns separately. Doing so would change the intended Aitchison metric. The four-cluster solution is a teaching resolution to compare methods, not a discovered number of natural basketball roles.

## Fuzzy c-means

Fuzzy c-means estimates centers $\mathbf v_1,\ldots,\mathbf v_K$ and memberships $u_{ik}$ by minimizing

$$
J_m(U,V)=\sum_{i=1}^n\sum_{k=1}^K u_{ik}^{m}\|\mathbf z_i-\mathbf v_k\|^2,
\qquad u_{ik}\ge0,\quad\sum_ku_{ik}=1,\quad m>1.
$$

The exponent $m$ controls fuzziness. Values near 1 favor hard assignments. Larger values distribute membership more evenly. For nonzero distances, the alternating updates are

$$
\mathbf v_k=\frac{\sum_i u_{ik}^{m}\mathbf z_i}{\sum_i u_{ik}^{m}},
\qquad
u_{ik}=\left[\sum_{\ell=1}^{K}
\left(\frac{\|\mathbf z_i-\mathbf v_k\|}{\|\mathbf z_i-\mathbf v_\ell\|}\right)^{2/(m-1)}\right]^{-1}.
$$

When a point exactly equals a center, the zero-distance case requires a separate assignment rule. The package handles this case. Like k-means, fuzzy c-means can reach different local solutions, so use multiple starts.

```{r}
set.seed(632)
fuzzy_fits <- lapply(1:20, function(b)
  e1071::cmeans(Z, centers = 4, m = 2, iter.max = 200))
fuzzy <- fuzzy_fits[[which.min(vapply(fuzzy_fits, function(f) f$withinerror, numeric(1)))]]
U <- fuzzy$membership
stopifnot(max(abs(rowSums(U) - 1)) < 1e-8)
membership_table <- data.frame(player = player_names, attempts = rowSums(counts),
  maximum_membership = apply(U, 1, max), U)
membership_table %>% arrange(maximum_membership) %>% slice_head(n = 8) %>%
  knitr::kable(digits = 3, row.names = FALSE)
```

Fuzzy memberships are weights in an optimization problem. They are not calibrated probabilities that a player has a particular latent role. By contrast, GMM responsibilities are conditional probabilities under a fitted generative model. Neither automatically captures uncertainty caused by having only a finite sample of shots.

```{r}
#| fig-cap: "Shot-selection profiles in two CLR principal coordinates. Transparency indicates maximum fuzzy membership."
#| fig-alt: "Player points colored by their largest fuzzy membership with uncertain boundary points shown more transparently."
display <- data.frame(player = player_names, PC1 = coordinates[, 1], PC2 = coordinates[, 2],
  cluster = factor(max.col(U)), certainty = apply(U, 1, max))
ggplot(display, aes(PC1, PC2, color = cluster, alpha = certainty)) +
  geom_point(size = 2) + scale_alpha_continuous(range = c(0.25, 1)) +
  coord_equal() + theme_minimal() + labs(alpha = "Largest fuzzy weight")
```

### Center interpretation on the original scale

For a CLR center $\mathbf v_k$, apply inverse closure: $p_{kz}=\exp(v_{kz})/\sum_v\exp(v_{kv})$. This gives a composition for the representative style. It is not generally the arithmetic average of member shares.

```{r}
center_shares <- exp(fuzzy$centers) / rowSums(exp(fuzzy$centers))
round(center_shares, 3) %>% knitr::kable()
```

Assign descriptive names only after examining these centers and member examples. Cluster labels such as 1 and 2 have no ordered meaning.

## Bayesian uncertainty in shot selection

Use a sampling model for the observed counts:

$$
\mathbf c_i\mid\mathbf p_i\sim\operatorname{Multinomial}(n_i,\mathbf p_i),
\qquad\mathbf p_i\sim\operatorname{Dirichlet}(\boldsymbol\alpha),
$$

so that

$$
\mathbf p_i\mid\mathbf c_i\sim\operatorname{Dirichlet}(\boldsymbol\alpha+\mathbf c_i).
$$

For each posterior draw of all player profiles, transform to CLR coordinates and refit k-means. Record whether every pair lands together. The **co-clustering matrix** is

$$
C_{ij}=\frac1B\sum_{b=1}^{B}\mathbb 1\{z_i^{(b)}=z_j^{(b)}\}.
$$

Because this uses pairs, arbitrary label permutations across runs do not matter. $C_{ij}$ summarizes the chosen clustering algorithm applied to posterior profile draws. It is not the posterior co-membership probability of a fitted Bayesian mixture model. Random starts can add algorithm variability, so we use multiple starts and keep $K$ fixed.

```{r}
draw_dirichlet_rows <- function(shape) {
  G <- matrix(rgamma(length(shape), shape = as.vector(shape)),
              nrow = nrow(shape), ncol = ncol(shape))
  G / rowSums(G)
}
set.seed(633)
B <- 100
C <- matrix(0, nrow(counts), nrow(counts))
for (b in seq_len(B)) {
  P_draw <- draw_dirichlet_rows(posterior_shape)
  labels <- kmeans(clr(P_draw), centers = 4, nstart = 20, iter.max = 100)$cluster
  C <- C + outer(labels, labels, "==") / B
}
stopifnot(max(abs(diag(C) - 1)) < 1e-10, max(abs(C - t(C))) < 1e-10)
consensus <- cutree(hclust(as.dist(1 - C), method = "average"), k = 4)
```

Average linkage accepts this pairwise dissimilarity. Avoid automatically applying Ward's squared-Euclidean interpretation to an arbitrary consensus dissimilarity.

```{r}
#| fig-cap: "Co-clustering frequencies from uncertain shot-selection profiles, ordered by consensus cluster."
#| fig-alt: "A square heat map with strong blocks for stable groups and lighter boundaries for uncertain pairs."
ordering <- order(consensus)
image(seq_len(nrow(C)), seq_len(nrow(C)), C[ordering, ordering],
  col = hcl.colors(30, "Blues"), zlim = c(0, 1),
  xlab = "Players ordered by consensus cluster", ylab = "Players in the same order",
  main = "Co-clustering frequency (0 to 1)")
```

The multinomial assumes a stable shot-selection vector and conditionally independent attempts. Game-by-game role changes violate this. A next step is resampling complete games within player or fitting a dynamic multinomial model. Measurement uncertainty, sampling stability, and genuine changes in style are different sources of variation.

## Sensitivity to preprocessing and fuzziness

```{r}
set.seed(634)
fuzziness <- bind_rows(lapply(c(1.3, 1.7, 2), function(m) {
  candidates <- lapply(1:10, function(b) e1071::cmeans(Z, centers = 4, m = m, iter.max = 200))
  f <- candidates[[which.min(vapply(candidates, function(f) f$withinerror, numeric(1)))]]
  data.frame(m = m, mean_maximum_membership = mean(apply(f$membership, 1, max)),
    ARI_vs_kmeans = mclust::adjustedRandIndex(max.col(f$membership), hard$cluster))
}))
fuzziness %>% knitr::kable(digits = 3)

smoothing <- bind_rows(lapply(c(0.1, 0.5, 2), function(a) {
  smoothed <- (counts + a) / (rowSums(counts) + ncol(counts) * a)
  fit <- kmeans(clr(smoothed), centers = 4, nstart = 50)
  data.frame(alpha_per_zone = a,
    ARI_vs_baseline = mclust::adjustedRandIndex(hard$cluster, fit$cluster))
}))
smoothing %>% knitr::kable(digits = 3)
```

Do not compare fuzzy objective values across different $m$ as if they were the same criterion. Interpret how the membership patterns change. Repeat the analysis over $K=3,4,5,6$ and compare descriptive usefulness, stability, and cluster sizes.

## A complementary resampling check

The posterior experiment holds the observed player roster fixed and varies its latent shot profiles. To test sensitivity to which players define the centers, sample 80% of players without replacement, fit centers, and assign **all** player profiles to those centers. This yields comparable partitions for ARI, including players omitted from the fit.

```{r}
nearest_center <- function(Z, centers) {
  distances <- sapply(seq_len(nrow(centers)), function(k)
    rowSums(sweep(Z, 2, centers[k, ], "-")^2))
  max.col(-distances, ties.method = "first")
}
set.seed(635)
resample_ari <- replicate(50, {
  rows <- sample(nrow(Z), floor(0.8 * nrow(Z)))
  fit <- kmeans(Z[rows, ], centers = 4, nstart = 20, iter.max = 100)
  assigned <- nearest_center(Z, fit$centers)
  mclust::adjustedRandIndex(hard$cluster, assigned)
})
quantile(resample_ari, c(0.1, 0.5, 0.9))
```

These two experiments answer different questions. Posterior profile draws test uncertainty about shot selection under a count model. Player subsampling tests sensitivity to the roster used to define the centers. Neither alone proves that natural discrete roles exist.

## Method selection for a sports project

| Need | Candidate approach | Main check |
|---|---|---|
| Compact standardized numeric profiles | k-means | Multiple starts and feature sensitivity |
| Actual player representatives or mixed features | PAM with Gower distance | Variable weights and missingness |
| A hierarchy of similarity | Hierarchical clustering | Linkage and cut sensitivity |
| Probabilistic elliptical components | GMM | Covariance regularization and model fit |
| Dense spatial regions and noise | DBSCAN or HDBSCAN | Density settings and noise fraction |
| Connected nonconvex groups | Spectral clustering | Graph construction and isolated nodes |
| Additive shot patterns | NMF | Rank and reconstruction stability |
| Graded archetype weights | Fuzzy c-means | Fuzzifier and center interpretation |
| Shot-allocation proportions | Log-ratio distance with clustering | Zero handling and compositional geometry |
| Uncertain inputs or unstable boundaries | Consensus across perturbations | The exact perturbation mechanism |

## Student task and discussion answers

1. Explain why $u_{i1}=0.6$ from fuzzy c-means is not automatically a 60% probability that player $i$ has role 1.
2. A pair has $C_{ij}=0.93$. Write a precise interpretation naming the data model and the clustering algorithm.
3. Explain why a three-share CLR vector has only two independent coordinates.
4. Submit a one-page scouting summary with the unit of analysis, eligibility rule, feature definitions, centers in original units, two ambiguous players, and a stability statement.

::: {.callout-tip collapse="true" title="Discussion answers"}
Fuzzy weights optimize a distance-based objective and depend on $m$, without a likelihood for latent roles. The pair clustered together in 93% of the k-means runs applied to Dirichlet posterior draws of shot-selection profiles, using four clusters and the stated initialization procedure. CLR coordinates sum to zero, so knowing two determines the third. The scouting summary should describe shooting style separately from quality and should avoid using cluster numbers as a ranking.
:::

## Further reading

- [e1071 manual, cmeans](https://cran.r-project.org/web/packages/e1071/e1071.pdf): fuzzy objective and R implementation.
- [HDBSCAN R documentation](https://search.r-project.org/CRAN/refmans/dbscan/html/hdbscan.html): density-based membership strength.
- [mclust documentation](https://mclust-org.github.io/mclust/): probabilistic mixtures and classification uncertainty.
- Aitchison, J. (1986), *The Statistical Analysis of Compositional Data*: log-ratio geometry.
