---
title: "5.1 Team Evaluation: Outcomes, Luck, and Schedules"
subtitle: "Why records mislead, and what a rating has to fix"
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
source("R/team_helpers.R")

nfl <- load_nfl_games()
epl <- load_epl_matches()
```

## Learning objectives

After this tutorial, you should be able to:

- define team strength as a latent quantity and list the observable outcomes that inform it;
- compare win percentage with point differential as measures of strength and explain the Pythagorean expectation;
- decompose the variance of win percentage into talent and luck and estimate the number of games at which they are equal;
- show why an unadjusted margin depends on schedule and why adjustment requires solving for every team at once; and
- describe the special features of soccer outcomes: low scores, draws, and home advantage.

## Strength is latent; outcomes are observed

Let $\theta_j$ denote the strength of team $j$: the expected margin against an average opponent on a neutral field. No game reveals $\theta_j$ directly. A game between teams $i$ and $j$ reveals a noisy function of $\theta_i-\theta_j$: a winner, a margin, or a pair of goal counts. Every rating system in this lecture is a rule for turning those outcomes into estimates of $\theta$, and every one of them has to deal with three facts.

1. **Outcomes are noisy.** A single NFL game has a margin standard deviation near 14 points around its expectation.
2. **Schedules differ.** A team's results depend on who it played.
3. **Home matters.** The home team wins more often than not in every major league.

## Win percentage and point differential

```{r}
standings_2023 <- team_season_table(nfl %>% filter(season == 2023)) %>%
  arrange(desc(win_pct))

standings_2023 %>%
  select(team, wins, ties, win_pct, points_for, points_against, point_diff) %>%
  slice_head(n = 8) %>%
  knitr::kable(digits = 3)
```

Win percentage throws away the size of every victory. Point differential keeps it. Which is the better measure of strength? One test is prediction: which one better forecasts the *next* season's record?

```{r}
standings <- team_season_table(nfl)

next_season <- standings %>%
  select(season, team, win_pct, point_diff) %>%
  inner_join(
    standings %>% transmute(season = season - 1, team, next_win_pct = win_pct),
    by = c("season", "team")
  )

next_season %>%
  summarize(
    seasons = paste(unique(season), collapse = ", "),
    cor_win_pct_with_next = cor(win_pct, next_win_pct),
    cor_point_diff_with_next = cor(point_diff, next_win_pct)
  ) %>%
  knitr::kable(digits = 3)
```

With only two season-to-season transitions the two correlations are statistically indistinguishable. Over long histories point differential has been the better predictor, which is the basis of the Pythagorean expectation below: a 30-point win and a 1-point win are different evidence about strength, and the record discards the difference.

### Pythagorean expectation

Bill James observed that a team's win percentage is close to

$$
\widehat{\mathrm{WP}}=\frac{\mathrm{PF}^{\,k}}{\mathrm{PF}^{\,k}+\mathrm{PA}^{\,k}},
$$

where $\mathrm{PF}$ and $\mathrm{PA}$ are points for and against. The exponent $k$ depends on the sport; for the NFL a common value is 2.37. The formula converts a point differential into an expected record, and the gap between actual and expected wins is a natural definition of luck.

```{r}
standings_2023 <- standings_2023 %>%
  mutate(
    pythagorean = pythagorean(points_for, points_against, exponent = 2.37),
    luck_wins = (win_pct - pythagorean) * games
  )

standings_2023 %>%
  arrange(desc(luck_wins)) %>%
  select(team, win_pct, pythagorean, luck_wins) %>%
  slice(c(1:4, 29:32)) %>%
  knitr::kable(digits = 3)
```

```{r}
#| fig-cap: "Actual win percentage against Pythagorean expectation for the 2023 NFL season."
#| fig-alt: "Scatter plot along the diagonal with team labels; a few teams sit noticeably above or below the line."
ggplot(standings_2023, aes(pythagorean, win_pct)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(color = "#03a9e6", size = 2) +
  ggrepel::geom_text_repel(aes(label = team), size = 2.8) +
  labs(x = "Pythagorean expectation", y = "Actual win percentage") +
  theme_minimal()
```

Teams above the line won more close games than their scoring suggests; teams below it lost more. Close-game records are notoriously unstable from season to season, which is why the Pythagorean record is the better predictor.

## Talent and luck

Treat each team's season win percentage as a noisy measurement of its true winning probability $p_j$:

$$
\widehat{\mathrm{WP}}_j=p_j+e_j,
\qquad
\operatorname{Var}(e_j)=\frac{p_j(1-p_j)}{G},
$$

where $G$ is the number of games. Across the league,

$$
\operatorname{Var}\!\left(\widehat{\mathrm{WP}}\right)=\underbrace{\operatorname{Var}(p)}_{\text{talent}}+\underbrace{\frac{\bar p(1-\bar p)}{G}}_{\text{luck}}.
$$

This is the signal-plus-noise decomposition of Lecture 4 applied to teams. The luck variance is known from the binomial, so the talent variance is the observed variance minus it.

```{r}
luck_variance <- 0.25 / 17
observed_variance <- var(standings$win_pct)
talent_variance <- observed_variance - luck_variance

data.frame(
  observed_variance = observed_variance,
  luck_variance = luck_variance,
  talent_variance = talent_variance,
  talent_share = talent_variance / observed_variance,
  games_for_equal_share = 0.25 / talent_variance
) %>%
  knitr::kable(digits = 4)
```

Roughly 40% of the spread in NFL records over these three seasons is luck, and talent and luck contribute equally at about twelve games, so a 17-game season is only modestly more signal than noise. The same reasoning gives the shrinkage estimate of a team's true winning probability,

$$
\hat p_j=\bar p+\frac{\operatorname{Var}(p)}{\operatorname{Var}(p)+\bar p(1-\bar p)/G}\left(\widehat{\mathrm{WP}}_j-\bar p\right),
$$

which regresses every record toward .500 by a factor equal to the talent share.

## Schedules

A team's average margin also depends on whom it played. Define the naive rating as the average margin,

$$
r_j^{(0)}=\frac{1}{G_j}\sum_{g\in\mathcal G_j}\left(\text{points for}-\text{points against}\right)_g.
$$

A fair rating should add back the quality of the opposition: a team that averages $+3$ against opponents who are themselves $+2$ is better than a team that averages $+3$ against opponents who are $-2$. That suggests the iteration

$$
r_j^{(t+1)}=\frac{1}{G_j}\sum_{g\in\mathcal G_j}\left[\text{margin}_g+r^{(t)}_{\text{opp}(g)}\right],
$$

which repeatedly adjusts each rating by the average current rating of its opponents.

```{r}
games_2023 <- nfl %>% filter(season == 2023)
teams <- sort(unique(games_2023$home_team))

long <- bind_rows(
  games_2023 %>% transmute(team = home_team, opponent = away_team, margin = margin),
  games_2023 %>% transmute(team = away_team, opponent = home_team, margin = -margin)
)

rating <- setNames(rep(0, length(teams)), teams)
for (iteration in seq_len(50)) {
  rating <- long %>%
    mutate(opponent_rating = rating[opponent]) %>%
    group_by(team) %>%
    summarize(rating = mean(margin + opponent_rating), .groups = "drop") %>%
    with(setNames(rating, team))
  rating <- rating - mean(rating)
}

schedule_adjusted <- long %>%
  group_by(team) %>%
  summarize(average_margin = mean(margin), .groups = "drop") %>%
  mutate(
    adjusted_rating = rating[team],
    strength_of_schedule = adjusted_rating - average_margin
  ) %>%
  arrange(desc(adjusted_rating))

schedule_adjusted %>%
  slice_head(n = 8) %>%
  knitr::kable(digits = 2)
```

The iteration converges to a fixed point in which every rating equals average margin plus average opponent rating. Tutorial 5.2 shows that this fixed point is the least-squares solution of a regression with one indicator column per team, which is a far more useful way to think about it because it extends to priors, uncertainty, and different outcomes.

## Soccer outcomes

Soccer changes the problem in three ways.

```{r}
epl %>%
  group_by(season) %>%
  summarize(
    matches = n(),
    home_goals = mean(home_goals),
    away_goals = mean(away_goals),
    home_win = mean(result == "H"),
    draw = mean(result == "D"),
    away_win = mean(result == "A"),
    .groups = "drop"
  ) %>%
  knitr::kable(digits = 3)
```

- **Low scores.** With about 1.5 home goals and 1.2 away goals per match, the margin takes only a handful of values, and a count model is more natural than a normal one.
- **Draws.** About a quarter of matches are drawn. A win-loss model must handle a third outcome.
- **Home advantage.** The home team scores about 0.3 goals more and wins about 45% of matches, which is a large effect in a low-scoring sport.

```{r}
#| fig-cap: "Distribution of goals per team in a match across three Premier League seasons, with the Poisson distribution at the same mean."
#| fig-alt: "Bar chart of goal counts from 0 to 7 with Poisson probabilities overlaid as points."
goal_counts <- c(epl$home_goals, epl$away_goals)
goal_table <- data.frame(goals = 0:7) %>%
  mutate(
    observed = sapply(goals, function(g) mean(goal_counts == g)),
    poisson = dpois(goals, mean(goal_counts))
  )

ggplot(goal_table, aes(goals)) +
  geom_col(aes(y = observed), fill = "#03a9e6", alpha = 0.7) +
  geom_point(aes(y = poisson), color = "black", size = 2) +
  labs(x = "Goals scored by one team", y = "Proportion of team-matches", caption = "Points: Poisson with the same mean") +
  theme_minimal()
```

The Poisson distribution fits goal counts well, which is the foundation of the model in Tutorial 5.4.

## Practice

1. A team goes 12-5 with a point differential of $+20$. Explain in one sentence why its Pythagorean record is far below 12-5.
2. Using the talent variance estimated above, what is the shrinkage estimate of true winning probability for a 13-4 team?
3. Why does the schedule-adjustment iteration need to recenter the ratings at every step?
4. Give one reason a normal model for margins is less appropriate for soccer than for the NFL.

::: {.callout-tip collapse="true" title="Show answers"}
1. A $+20$ differential over 17 games is about $+1.2$ per game, which corresponds to only slightly better than a .500 team; the 12-5 record came from winning close games.
2. $\hat p=0.5+s(13/17-0.5)$ where $s$ is the talent share; with $s\approx0.58$ the estimate is about .65.
3. Adding the average opponent rating to every team's rating shifts the whole set by a constant, so without recentering the ratings would drift; only differences between ratings are identified.
4. Margins in soccer are small integers with many zeros, so a continuous symmetric distribution cannot represent draws or the discreteness.
:::

## Key takeaways

- Team strength is latent; outcomes are noisy functions of strength differences plus home advantage.
- Point differential carries more information than the record and predicts future records at least as well.
- About half the spread in NFL records over one season is luck; shrink records toward .500 accordingly.
- Schedule adjustment is a system of equations, one per team, which is a regression in disguise.
- Soccer needs count models and a third outcome.
