---
title: "5.6 Evaluating and Communicating Team Ratings"
subtitle: "Honest forecast scoring, the betting market as a benchmark, and rankings with uncertainty"
---

```{r}
#| label: setup
#| include: false
library(dplyr)
library(ggplot2)
source("R/team_helpers.R")

nfl <- load_nfl_games()
state_space <- readRDS("data/team_results.rds")$state_space
```

## Learning objectives

After this tutorial, you should be able to:

- evaluate sequential forecasts without look-ahead and explain why a warm-up period is needed;
- convert moneylines and spreads into probabilities and remove the bookmaker's margin;
- compare rating systems with the market on log loss, Brier score, and calibration;
- explain break-even probabilities, expected value, and why beating the market is hard;
- turn a posterior over ratings into probabilities for rankings; and
- write a rating report that states its scale, uncertainty, and limitations.

## The rule for scoring sequential forecasts

A forecast for a game may use only information available before kickoff. For Elo and the Kalman filter this is automatic, because both produce pregame ratings. For a static regression it means refitting before every week on the games already played. The 2022 season serves as a warm-up so that every system has separated the teams before scoring starts in 2023; ties are excluded because a home-win probability has no natural score for them.

```{r}
hyper <- state_space$hyper_summary
get <- function(name) hyper$mean[hyper$variable == name]

elo <- run_elo(nfl, k = 20, home_advantage = 55, margin_multiplier = TRUE)
kalman <- kalman_ratings(
  nfl,
  sigma_obs = get("sigma_obs"), sigma_week = get("sigma_week"),
  sigma_season = get("sigma_season"), rho = get("rho"),
  home_advantage = get("h"), initial_sd = get("sigma_initial")
)

# Rolling ridge: before each week, refit on the current season's earlier games.
rolling_ridge <- nfl %>%
  mutate(ridge_margin = NA_real_)
for (s in unique(nfl$season)) {
  for (w in sort(unique(nfl$week[nfl$season == s]))) {
    rows <- which(nfl$season == s & nfl$week == w)
    history <- nfl %>% filter(season == s, week < w)
    if (nrow(history) < 16) {
      rolling_ridge$ridge_margin[rows] <- 2 * (1 - nfl$neutral_site[rows])
    } else {
      fit <- ridge_ratings(history, lambda = 7)
      rolling_ridge$ridge_margin[rows] <- fit$home_advantage * (1 - nfl$neutral_site[rows]) +
        fit$ratings[nfl$home_team[rows]] - fit$ratings[nfl$away_team[rows]]
    }
  }
}

forecasts <- nfl %>%
  select(game_id, season, week, home_team, away_team, margin, home_win, tie, spread_line, home_moneyline, away_moneyline) %>%
  left_join(elo$games %>% select(game_id, elo_prob = home_prob), by = "game_id") %>%
  left_join(kalman$predictions %>% select(game_id, kalman_margin = predicted_margin, kalman_prob = home_prob), by = "game_id") %>%
  left_join(rolling_ridge %>% select(game_id, ridge_margin), by = "game_id") %>%
  mutate(
    ridge_prob = pnorm(ridge_margin / 13),
    market_prob = remove_vig(moneyline_to_probability(home_moneyline), moneyline_to_probability(away_moneyline))[, 1],
    spread_prob = pnorm(spread_line / 13.5)
  ) %>%
  filter(season >= 2023, tie == 0, !is.na(market_prob))

nrow(forecasts)
```

## The market as a benchmark

A moneyline of $-150$ means a bettor risks 150 to win 100; a moneyline of $+130$ means risking 100 wins 130. The implied probabilities are

$$
p_{\text{implied}}=
\begin{cases}
\dfrac{|M|}{|M|+100}, & M<0,\\[10pt]
\dfrac{100}{M+100}, & M>0.
\end{cases}
$$

The two sides of a game imply probabilities that sum to more than one. The excess is the bookmaker's margin, called the **vig** or **hold**, and dividing each implied probability by the sum removes it.

```{r}
example <- nfl %>% filter(season == 2023, week == 1) %>% slice_head(n = 4)
example %>%
  transmute(
    home_team, away_team, home_moneyline, away_moneyline,
    implied_home = moneyline_to_probability(home_moneyline),
    implied_away = moneyline_to_probability(away_moneyline),
    hold = implied_home + implied_away - 1,
    fair_home = remove_vig(implied_home, implied_away)[, 1]
  ) %>%
  knitr::kable(digits = 3)
```

The closing spread is a market forecast of the margin, and $\Phi(\text{spread}/13.5)$ turns it into a probability using the historical noise in NFL margins.

## Comparing the systems

```{r}
score <- function(probability, label) {
  data.frame(
    model = label,
    log_loss = log_loss(forecasts$home_win, probability),
    brier = brier_score(forecasts$home_win, probability),
    accuracy = mean((probability > 0.5) == forecasts$home_win)
  )
}

bind_rows(
  score(rep(mean(forecasts$home_win), nrow(forecasts)), "constant home-win rate"),
  score(forecasts$ridge_prob, "rolling ridge (within season)"),
  score(forecasts$elo_prob, "Elo with margin multiplier (K = 20)"),
  score(forecasts$kalman_prob, "Kalman filter (state-space)"),
  score(forecasts$spread_prob, "market: closing spread"),
  score(forecasts$market_prob, "market: moneyline, vig removed")
) %>%
  knitr::kable(digits = 4)
```

```{r}
data.frame(
  forecast = c("Kalman filter", "rolling ridge", "closing spread"),
  rmse_margin = c(
    sqrt(mean((forecasts$margin - forecasts$kalman_margin)^2)),
    sqrt(mean((forecasts$margin - forecasts$ridge_margin)^2)),
    sqrt(mean((forecasts$margin - forecasts$spread_line)^2))
  )
) %>%
  knitr::kable(digits = 2)
```

Three patterns hold across seasons. The market is the best forecaster; it incorporates injuries, weather, and information that results alone do not contain. The state-space model, which carries information across seasons and weights it correctly, is the best of the results-only systems. The rolling ridge, which starts every season from nothing, is the weakest, especially in early weeks.

### Calibration

```{r}
#| fig-cap: "Calibration of three forecasters on 2023 and 2024 games. Points are bins of predicted probability; the horizontal position is the mean forecast and the vertical position the observed home-win rate."
#| fig-alt: "Three sets of points near the diagonal, with more scatter for the results-only models."
calibration <- bind_rows(
  forecasts %>% transmute(model = "Elo", probability = elo_prob, home_win),
  forecasts %>% transmute(model = "Kalman filter", probability = kalman_prob, home_win),
  forecasts %>% transmute(model = "market (moneyline)", probability = market_prob, home_win)
) %>%
  mutate(bin = cut(probability, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE)) %>%
  group_by(model, bin) %>%
  summarize(mean_forecast = mean(probability), observed = mean(home_win), games = n(), .groups = "drop") %>%
  filter(games >= 10)

ggplot(calibration, aes(mean_forecast, observed, color = model)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(aes(size = games)) +
  geom_line() +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "Mean forecast probability", y = "Observed home-win rate", color = NULL, size = "Games") +
  theme_minimal()
```

With roughly 540 games, each bin holds a few dozen games and the observed rates bounce around the diagonal by sampling noise alone. None of the forecasters shows a systematic pattern of over- or underconfidence.

## Break-even, expected value, and the difficulty of beating the market

A bet at decimal odds $o$ (American $+M$ converts to $o=1+M/100$; $-M$ converts to $o=1+100/M$) returns $o$ per unit staked if it wins. If the true probability is $p$, the expected profit per unit is

$$
\text{EV}=p\,o-1,
$$

and the bet breaks even at $p^{*}=1/o$, which is exactly the implied probability before the vig is removed. To profit, a model's probability must exceed the implied probability of the *bettor's* side, which already includes the hold, and it must do so reliably.

```{r}
edges <- forecasts %>%
  mutate(
    odds_home = ifelse(home_moneyline > 0, 1 + home_moneyline / 100, 1 + 100 / abs(home_moneyline)),
    odds_away = ifelse(away_moneyline > 0, 1 + away_moneyline / 100, 1 + 100 / abs(away_moneyline)),
    ev_home = kalman_prob * odds_home - 1,
    ev_away = (1 - kalman_prob) * odds_away - 1,
    bet = case_when(ev_home > 0.05 ~ "home", ev_away > 0.05 ~ "away", TRUE ~ "none"),
    profit = case_when(
      bet == "home" ~ ifelse(home_win == 1, odds_home - 1, -1),
      bet == "away" ~ ifelse(home_win == 0, odds_away - 1, -1),
      TRUE ~ 0
    )
  )

edges %>%
  filter(bet != "none") %>%
  summarize(
    bets = n(),
    win_rate = mean(profit > 0),
    total_profit_per_unit_staked = sum(profit) / n(),
    mean_model_edge = mean(pmax(ev_home, ev_away))
  ) %>%
  knitr::kable(digits = 3)
```

The Kalman model finds apparent edges in a large number of games, and the realized return on those bets is clearly negative. The edges were errors in the model, not in the market. That is the expected result for a results-only model against a market that knows more, and it is a useful calibration of ambition: a model that beats the market's log loss would be remarkable, and none of the systems here does.

::: {.callout-warning}
## Two cautions

First, an apparent edge on one season of games is within noise; a return of a few percent on a few hundred bets has a standard error of several percent. Second, the Kelly criterion $f^{*}=(po-1)/(o-1)$ gives the stake fraction that maximizes long-run growth *if* $p$ is correct; with an overconfident model it produces ruin faster than flat betting.
:::

## Rankings with uncertainty

The Stan posterior for the final period turns every question about the ranking into a probability.

```{r}
draws <- state_space$final_theta_draws
ranks <- t(apply(-draws, 1, rank, ties.method = "first"))
colnames(ranks) <- colnames(draws)

ranking <- data.frame(
  team = colnames(draws),
  posterior_mean = colMeans(draws),
  posterior_sd = apply(draws, 2, sd),
  probability_best = colMeans(ranks == 1),
  probability_top_5 = colMeans(ranks <= 5),
  expected_rank = colMeans(ranks)
) %>%
  arrange(desc(posterior_mean))

ranking %>% slice_head(n = 10) %>% knitr::kable(digits = 2, row.names = FALSE)
```

```{r}
#| fig-cap: "Posterior rank distributions for the six strongest teams at the end of the 2024 regular season."
#| fig-alt: "Six histograms of rank; the top team's mass is concentrated near rank 1 but spreads to rank 5 or beyond."
top_teams <- ranking$team[1:6]

as.data.frame(ranks[, top_teams]) %>%
  tidyr::pivot_longer(everything(), names_to = "team", values_to = "rank") %>%
  mutate(team = factor(team, levels = top_teams)) %>%
  ggplot(aes(rank)) +
  geom_bar(fill = "#03a9e6") +
  facet_wrap(~team, ncol = 3) +
  coord_cartesian(xlim = c(1, 12)) +
  labs(x = "Rank", y = "Posterior draws") +
  theme_minimal()
```

Even the strongest team at season's end has a meaningful chance of being only the third- or fourth-best team. A published ranking that omits this is stating a posterior mode as if it were a fact.

## Communicating a rating

A rating report should let a reader answer four questions without contacting the analyst.

1. **What is the scale?** Points against an average opponent on a neutral field, log-odds, or Elo points. State it, and give the home advantage on the same scale.
2. **How uncertain is each rating?** Report an interval, or at least a standard deviation, and say what it excludes.
3. **When is the rating for?** A season-long average, a smoothed estimate for a date, or a forecast for the next game are different objects.
4. **What does the model not see?** Injuries, suspensions, weather, and motivation are absent from results-only models. Say so, and compare with the market where possible.

A reporting checklist for team ratings:

- outcome modeled (result, margin, or goals) and data window;
- design of the model, including home advantage, neutral sites, and ties;
- prior or penalty and the estimated variance components;
- for dynamic models, the drift and between-season assumptions;
- forecast evaluation on games not used for fitting, against a market benchmark;
- rankings as probabilities; and
- known limitations.

## Practice

1. Convert a moneyline of $-240$ to an implied probability and to decimal odds.
2. Two sides of a game have implied probabilities 0.58 and 0.46. What is the hold, and what are the fair probabilities?
3. Why is the rolling ridge model at its worst in weeks 2 through 4?
4. A model gives a team a 60% chance of winning at decimal odds of 1.80. What is the expected value per unit and the Kelly fraction?

::: {.callout-tip collapse="true" title="Show answers"}
1. $240/340\approx0.706$; decimal odds $1+100/240\approx1.417$.
2. Hold $=0.04$; fair probabilities $0.58/1.04\approx0.558$ and $0.46/1.04\approx0.442$.
3. It has seen only a handful of games per team and no prior information from the previous season, so its ratings are mostly noise and shrinkage pulls them all toward zero.
4. $\text{EV}=0.6\times1.8-1=0.08$; Kelly fraction $0.08/0.8=0.10$.
:::

## Key takeaways

- Score forecasts made only with pregame information, after a warm-up period, and exclude outcomes the forecast cannot express.
- Moneylines convert to probabilities; remove the hold before comparing with a model.
- The market beats results-only models; the state-space model is the best of those models and the rolling static model the worst.
- Apparent betting edges from a results-only model do not survive the hold; treat market log loss as the bar to clear.
- Report rankings as rank probabilities and state scale, uncertainty, timing, and blind spots.
